---
title: "Peer Assignment 2"
author: "Ahmed Tadde"
date: "June 18, 2015"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    keep_md: yes
---

##Synopsis

The U.S. National Oceanic and Atmospheric Administration (NOAA) maintains a database that tracks weather events in the United States. The data used for this analysis was collected from 1950 to 2011. The goal of the ensuing analysis is to:

   1. Determine the types of events (as indicated in the EVTYPE variable) that are most harmful with respect to population health(injuries and fatalities).
   2. Determine the types of events that have the greatest economic consequences( Crop and property Damage).

 

##Data Processing[Code Section]

   - Libraries
   - Loading the Data
   - Tidying the Data

###Libraries
```{r, tidy=TRUE}
library(data.table)
library(dplyr)
```




###Loading the Data
```{r, tidy = TRUE,cache=TRUE}


link <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
location.name <- "repdata-data-StormData.csv.bz2"
if(!file.exists("repdata-data-StormData.csv.bz2")) {
    download.file(link, destfile = location.name, method = "auto")
}

Data <- read.table("repdata-data-StormData.csv.bz2", 
                 sep = ",",
                 na.strings = "NA",
                 stringsAsFactors = F,
                 header = T)

Data <- as.data.table(Data)

#=====================================================================================
####Inspecting the Data
dim(Data)
str(Data)
names(Data)
```




###Tidying the Data
```{r, tidy = TRUE, cache=TRUE}
#=====================================================================================
#####We can set some variables to numeric type for calculations later on
Data$FATALITIES <- as.numeric(Data$FATALITIES)
Data$INJURIES <- as.numeric(Data$INJURIES)
# Data$CROPDMG <- as.numeric(Data$CROPDMG)
# Data$PROPDMG <- as.numeric(Data$PROPDMG)


#=====================================================================================
##### Setting the string characters in EVTYPE to lower cases for convenience
Data$EVTYPE <- with(Data, tolower(EVTYPE))






#=====================================================================================
##### Recoding the Property damage variable based on a mapping method referenced from NOAA documentation  
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("","-","?", "NA")] <- 0
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("+")] <- 1
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("0","1","2","3","4","5","6","7","8","9")] <- 10
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("H","h")] <- 100
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("K","k")] <- 1000
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("M","m")] <- 1000000
Data$PROPDMGEXP[Data$PROPDMGEXP %in% c("B","b")] <- 1000000000
Data$PROPDMGEXP <- as.numeric(Data$PROPDMGEXP)





#=====================================================================================
##### Recoding the Crop damage variable based on a mapping method referenced from NOAA documentation
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("","-","?","NA")] <- 0
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("+")] <- 1
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("0","1","2","3","4","5","6","7","8","9")] <- 10
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("H","h")] <- 100
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("K","k")] <- 1000
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("M","m")] <- 1000000
Data$CROPDMGEXP[Data$CROPDMGEXP %in% c("B","b")] <- 1000000000
Data$CROPDMGEXP <- as.numeric(Data$CROPDMGEXP)


#=====================================================================================
##### To have observations with complete case (all 48 official weather 
##### event types), we filter the data to keep only the observations starting from 1996. 
Data<- filter(Data, as.POSIXct(Data$BGN_DATE, format="%m/%d/%Y") > as.POSIXct("12/31/1995", format="%m/%d/%Y"))


#=====================================================================================
##### Only keeping the columns of interests
Data <- Data%>%select(c(8,23:28))


#=====================================================================================
##### Do not include "summary" observations
Data <- Data[!(grepl("summary",Data$EVTYPE)),]


#=====================================================================================
##### Only keep events that created at least one type of damage.
Data <- Data%>%filter(
                      !(FATALITIES > 0.0 & INJURIES > 0.0 & 
                        PROPDMG > 0.0    & CROPDMG > 0.0)
                     )



#=====================================================================================
##### Create new variables to evaluate the harmful effect on 
##### the population (injuries+fatalities), the economic cost of
##### crop/property damages( separately and together) 
Data <- Data%>%mutate(
                      POPULATION.HEALTH = FATALITIES + INJURIES,
                      CROP.DAMAGE.COST  = CROPDMG * CROPDMGEXP,
                      PROPERTY.DAMAGE.COST = PROPDMG * PROPDMGEXP,
                      TOTAL.DAMAGE.COST = CROP.DAMAGE.COST + PROPERTY.DAMAGE.COST
                      )




#=====================================================================================
##### Mapping __some__ unofficial event types to the 48 event types [time consuming]
Data$EVTYPE[grepl("tstm wind",Data$EVTYPE) |
            grepl("severe thunderstorm",Data$EVTYPE)|
            grepl("severe thunderstorms",Data$EVTYPE)|
            grepl("thunderstorm",Data$EVTYPE)|
            grepl("thunderstorms",Data$EVTYPE)|
            grepl("thunderstorm winds",Data$EVTYPE)|
            grepl("thunderstorm wind",Data$EVTYPE)] <- "thunderstorm"


Data$EVTYPE[grepl("ice", Data$EVTYPE)|
            grepl("low temperature", Data$EVTYPE)|grepl("cool", Data$EVTYPE)| 
            grepl("icy road", Data$EVTYPE)|
            grepl("hypothermia/exposure", Data$EVTYPE)| 
            grepl("wintry mix", Data$EVTYPE)|
            grepl("snow", Data$EVTYPE)| 
            grepl("record low", Data$EVTYPE)|grepl("hypothermia", Data$EVTYPE)|
            grepl("record cold", Data$EVTYPE)|grepl("record  cold", Data$EVTYPE)|
            grepl("winter weather/mix", Data$EVTYPE)| 
            grepl("winter weather mix",Data$EVTYPE)] <- "winter weather"

Data$EVTYPE[grepl("frost",Data$EVTYPE)| grepl("freeze", Data$EVTYPE)] <- "frost/freeze"
Data$EVTYPE[grepl("heat", Data$EVTYPE) | grepl("hot", Data$EVTYPE)] <- "heat"
Data$EVTYPE[grepl("freezing", Data$EVTYPE)] <- "freezing fog"
Data$EVTYPE[grepl("marine tstm wind", Data$EVTYPE)] <- "marine thunderstorm wind"
Data$EVTYPE[grepl("hurricane", Data$EVTYPE) | grepl("typhoon", Data$EVTYPE)] <- "hurricane(typhoon)"
Data$EVTYPE[grepl("tornado", Data$EVTYPE)] <- "tornado"
Data$EVTYPE[grepl("surf", Data$EVTYPE)] <- "high surf"
Data$EVTYPE[grepl("cold/wind chill", Data$EVTYPE)] <- "cold/wind chill"
Data$EVTYPE[grepl("cold wind chill", Data$EVTYPE)] <- "cold/wind chill"
Data$EVTYPE[grepl("extreme cold wind chill", Data$EVTYPE)] <- "extreme cold/wind chill"
Data$EVTYPE[grepl("extreme cold/wind chill", Data$EVTYPE)] <- "extreme cold/wind chill"
Data$EVTYPE[grepl("hail", Data$EVTYPE)] <- "hail"
Data$EVTYPE[grepl("high winds", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("winds", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("wind and wave", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("wind", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("wind damage", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("whirl wind damage", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("strong wind", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("strong winds", Data$EVTYPE)] <-"high wind"
Data$EVTYPE[grepl("storm surge/tide", Data$EVTYPE)] <-"storm surge"
Data$EVTYPE[grepl("wild/forest", Data$EVTYPE)] <-"wildfire"
Data$EVTYPE[grepl("wild", Data$EVTYPE)] <- "wildfire"
Data$EVTYPE[grepl("fire", Data$EVTYPE)] <- "wildfire"
Data$EVTYPE[grepl("flood", Data$EVTYPE)| 
            grepl("hightide", Data$EVTYPE) | 
            grepl("high tide", Data$EVTYPE)|
            grepl("urban sml stream fld", Data$EVTYPE)|
            grepl("urban/sml stream fld", Data$EVTYPE)] <- "flood"


#=====================================================================================
####### Making the EVTYPE variable to a factor variable to facilitate calculations 
####### in the Result session
Data$EVTYPE <- as.factor(Data$EVTYPE)


```






##Results

- __The top 10 most harmful event types with respect to Injuries and with respect to Fatalities__ 

With respect to Injuries,
```{r, tidy = TRUE, cache = TRUE}
table1 <- Data[, .(INJURIES = sum(INJURIES, na.rm = T)),
              by = .(EVENT.TYPE = EVTYPE )]%>%arrange(desc(INJURIES))

table1 <- head(table1, n = 10L) ; table1

```


With respect to Fatalities,
```{r, tidy = TRUE , cache = TRUE}
table2 <- Data[, .(FATALITIES = sum(FATALITIES, na.rm = T)),
              by = .(EVENT.TYPE = EVTYPE )]%>%arrange(desc(FATALITIES))

table2 <- head(table2, n = 10L) ; table2

```


Displaying both results,
```{r, tidy = TRUE}
par(mfrow=c(1,2))
barplot(table1[1:10, INJURIES],
        names.arg=table1[1:10, EVENT.TYPE],
        horiz= T,  
        col="red", 
        xlab="Injuries", 
        main="Top 10 Most Costly Events (Population: Injuries)",
        xlim=c(0,25000),las=1,cex.names=.7, cex.main=0.8, cex.axis=1)

barplot(table2[1:10, FATALITIES],
        names.arg=table2[1:10, EVENT.TYPE],
        horiz= T,  
        col="red", 
        xlab="Fatalities", 
        main="Top 10 Most Costly Events (Population: Fatalities)",
        xlim=c(0,2500),las=1,cex.names=.7, cex.main=0.8, cex.axis=1)
```


- __The top 10 most harmful event types with respect to Crop Damage and with respect to Property Damage__ 


With respect to Crop Damage,
```{r, tidy = TRUE, cache = TRUE}

table3 <- Data[, .(CROP.DAMAGE.COST = sum(CROP.DAMAGE.COST, na.rm = T)/10^6),
              by = .(EVENT.TYPE = EVTYPE )]%>%arrange(desc(CROP.DAMAGE.COST))
table3 <- head(table3, n = 10L) ; table3

```


With respect to Property Damage,
```{r, tidy = TRUE, cache = TRUE}

table4 <- Data[, .(PROPERTY.DAMAGE.COST = sum(PROPERTY.DAMAGE.COST, na.rm = T)/10^6),
              by = .(EVENT.TYPE = EVTYPE )]%>%arrange(desc(PROPERTY.DAMAGE.COST))
table4 <- head(table4, n = 10L) ; table4

```


Displaying both results,
```{r, tidy = TRUE}
par(mfrow=c(1,2))
barplot(table3[1:10,CROP.DAMAGE.COST],
        names.arg=table3[1:10, EVENT.TYPE],
        horiz= T,  
        col="dark blue", 
        xlab="Dollars (x1,000,000)", 
        main="Top 10 Most Costly Events (Economically: Crop Damage)",
        xlim=c(0,20000),las=1,cex.names=.65, cex.main=0.8, cex.axis=1)

barplot(table4[1:10,PROPERTY.DAMAGE.COST],
        names.arg=table4[1:10, EVENT.TYPE],
        horiz= T,  
        col="dark blue", 
        xlab="Dollars (x1,000,000)", 
        main="Top 10 Most Costly Events (Economically: Property Damage)",
        xlim=c(0,200000),las=1,cex.names=.65, cex.main=0.8, cex.axis=1)

```



 -__The top 10 event types that are most harmful with respect to population health(injuries and fatalities) and economic consequences__


With respect to population health(injuries and fatalities),
```{r, tidy = TRUE}
table5 <- Data[, .(POPULATION.HEALTH = sum(POPULATION.HEALTH, na.rm = T)),
               by = .(EVENT.TYPE = EVTYPE )]%>%arrange(desc(POPULATION.HEALTH))
table5 <- head(table5, n = 10L) ; table5

```


With respect to economic consequences,
```{r, tidy = TRUE}
table6 <- Data[, .(TOTAL.DAMAGE.COST = sum(TOTAL.DAMAGE.COST, na.rm = T)/10^6),
               by = .(EVENT.TYPE = EVTYPE )]%>%arrange(desc(TOTAL.DAMAGE.COST))
table6 <- head(table6, n = 10L) ; table6


```


Displaying both results,
```{r, tidy = TRUE}
par(mfrow = c(1,2))
barplot(table5[1:10, POPULATION.HEALTH],
        names.arg=table5[1:10, EVENT.TYPE],
        horiz= T,  
        col="red", 
        xlab="Injuries and Fatalities", 
        main="Top 10 Most Costly Events (Population: Injuries and Fatalities)",
        xlim=c(0,25000),las=1,cex.names=.7, cex.main=0.8, cex.axis=1)



barplot(table6[1:10,TOTAL.DAMAGE.COST],
        names.arg=table2[1:10, EVENT.TYPE],
        horiz= T,  
        col="dark blue", 
        xlab="Dollars (x1,000,000)", 
        main="Top 10 Most Costly Events (Economically: Crop and Property)",
        xlim=c(0,250000),las=1,cex.names=.8, cex.main=0.8, cex.axis=1)

```



##Conlcusion

- In no particular order, excessive heat, tornadoes, and floods are the three event types that cause the most damage in terms of population health and in terms economict cost.


##References

- [NOAA Documentation](https://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf)
- [The 48 official event types since 1996](http://www.ncdc.noaa.gov/stormevents/details.jsp)